Generalized spaces for constructive algebra

### 2020-04-05 paper-herrsching2019/main.tex
Simplification by suppressing variables

However, the more powerful version is not constructive. This puts us in the
curious situation that we are promised, unconstructively, that there exists
a constructive proof. The textbook argument~\cite[Theorem~7.57]{johnstone:topos-theory} establishing the more powerful
version of Barr's theorem requires
the axiom of choice, but recently Michael Rathjen showed that~\textsc{zf}
suffices~\cite{rathjen:barr}.

@@

To show that~(3) implies~(2), we combine the double negation translation
with Friedman's trick. We give a sketch of the main thrust of the argument,
omitting lengthy verifications. Let~$\sigma \equiv (\alpha \seq{\vec x} \beta)$.
We introduce the operator~$\nabla$ with~$\nabla\varphi \defeqv
((\varphi \Rightarrow \beta) \Rightarrow \beta)$, where care has to be taken to
avoid capturing the variables~$\vec x$. This operator is a \emph{local
operator} in the sense that for any formula~$\varphi$ in some context~$\vec y$,
infinitary intuitionistic logic proves the sequents
\[
\top \seq{\vec x,\vec y} (\varphi \Rightarrow \nabla\varphi),\
\top \seq{\vec x,\vec y} (\nabla\nabla\varphi \Rightarrow \nabla\varphi),\
\top \seq{\vec x,\vec y} (\nabla\varphi \wedge \nabla\psi \Leftrightarrow \nabla(\varphi \wedge \psi)).
\]
We then set up the~\emph{$\nabla$-translation}~$\varphi \mapsto \varphi^\nabla$
for infinitary first-order formulas over the signature of~$\TT$ as described in
Table~\ref{table:nabla-translation}.
Again, care has to be taken to ensure that there is no variable collision with the
variables~$\vec x$. The chosen local operator is special in that
the~$\nabla$-translation of the law of excluded middle,
\[ \top \seq{\vec x,\vec y} \nabla(\varphi^\nabla \vee (\varphi^\nabla
\Rightarrow \nabla\bot)) \]
is intuitionistically provable. Straightforward inductions on the structure of
formulas or derivations then establish the following properties.
\begin{alphlist}
\item For any infinitary first-order formula~$\varphi$ in some context~$\vec y$,
infinitary intuitionistic logic proves~$\nabla(\varphi^\nabla) \seq{\vec x,\vec
y} \varphi^\nabla$.
\item For any geometric formula~$\varphi$ in some context~$\vec y$, infinitary
intuitionistic logic proves~$\varphi^\nabla \mathrel{\dashv\vdash\!\!\!_{\vec x,\vec y}} \nabla\varphi$.
\item For any derivation of a sequent~$\varphi \seq{\vec y} \psi$ in infinitary
classical logic, there is a derivation of~$\varphi^\nabla \seq{\vec y}
\psi^\nabla$ in infinitary intuitionistic logic.
\end{alphlist}
Combining these ingredients, the given derivation of~$\sigma \equiv (\alpha
\seq{\vec x} \beta)$ in infinitary classical logic can be transformed to give a derivation of
\begin{align*}
  \alpha[\vec x'/\vec x] &\seq{\vec x,\vec x'}
  \nabla(\alpha[\vec x'/\vec x]) \seq{\vec x,\vec x'}
  (\alpha[\vec x'/\vec x])^\nabla \\
  &\seq{\vec x,\vec x'}
  (\beta[\vec x'/\vec x])^\nabla \seq{\vec x,\vec x'}
  \nabla(\beta[\vec x'/\vec x]) \mathrel{\dashv\vdash\!\!\!_{\vec x,\vec x'}}
  ((\beta[\vec x'/\vec x] \Rightarrow \beta) \Rightarrow \beta).
\end{align*}
in infinitary intuitionistic logic. Substituting~$\vec x'$ by~$\vec x$, we obtain
\[ \alpha \seq{\vec x} ((\beta \Rightarrow \beta) \Rightarrow \beta) \seq{\vec
x} \beta. \]

To verify that~(2) implies~(1), XXX
\end{proof}

@@

>>>>

However, the more powerful version is not constructive. This puts us in the
curious situation that we are promised, unconstructively, that there exist
certain constructive proofs. The textbook
argument~\cite[Theorem~7.57]{johnstone:topos-theory} establishing the more
powerful version of Barr's theorem requires the axiom of choice, but recently
Michael Rathjen lowered the required metatheory to~\textsc{zf}~\cite[Remark~4.2]{rathjen:barr}.

@@

To show that~(3) implies~(2), we combine the double negation translation
with Friedman's trick. We give a sketch of the main thrust of the argument,
omitting lengthy verifications. Let~$\sigma \equiv (\alpha \seq{\vec a} \beta)$.
For notational simplicity, we assume that the context~$\vec a$ is empty, though
all of the following can be adapted to the nonempty case if care is taken to
avoid variable capture.

We introduce the operator~$\nabla$ with~$\nabla\varphi \defeqv
((\varphi \Rightarrow \beta) \Rightarrow \beta)$. This operator is a
\emph{local operator}~\cite[Section~14.5]{goldblatt:topoi} in the sense that
for any formula~$\varphi$ in some context~$\vec x$, infinitary intuitionistic
logic proves the sequents
\[
\top \seq{\vec x} (\varphi \Rightarrow \nabla\varphi),\
\top \seq{\vec x} (\nabla\nabla\varphi \Rightarrow \nabla\varphi),\
\top \seq{\vec x} (\nabla\varphi \wedge \nabla\psi \Leftrightarrow \nabla(\varphi \wedge \psi)).
\]
We then set up the~\emph{$\nabla$-translation}~$\varphi \mapsto \varphi^\nabla$
for infinitary first-order formulas over the signature of~$\TT$ as described in
Table~\ref{table:nabla-translation}.
Again, care has to be taken to ensure that there is no variable collision with the
variables~$\vec x$. The chosen local operator is special in that
the~$\nabla$-translation of the law of excluded middle,
\[ \top \seq{\vec x} \nabla(\varphi^\nabla \vee (\varphi^\nabla \Rightarrow \nabla\bot)), \]
is intuitionistically provable. Straightforward inductions on the structure of
formulas or derivations then establish the following properties.
\begin{alphlist}
\item For any infinitary first-order formula~$\varphi$ in some context~$\vec y$,
infinitary intuitionistic logic proves~$\nabla(\varphi^\nabla) \seq{\vec x} \varphi^\nabla$.
\item For any geometric formula~$\varphi$ in some context~$\vec x$, infinitary
intuitionistic logic proves~$\varphi^\nabla \mathrel{\dashv\vdash\!\!\!_{\vec x}} \nabla\varphi$.
\item For any derivation of a sequent~$\varphi \seq{\vec x} \psi$ in infinitary
classical logic, there is a derivation of~$\varphi^\nabla \seq{\vec x}
\psi^\nabla$ in infinitary intuitionistic logic.
\end{alphlist}
Combining these ingredients, the given derivation of~$\sigma \equiv (\alpha
\vdash \beta)$ in infinitary classical logic can be transformed to give a derivation of
\[
  \alpha \vdash
  \nabla\alpha \vdash
  \alpha^\nabla \vdash
  \beta^\nabla \vdash
  \nabla\beta \equiv
  ((\beta \Rightarrow \beta) \Rightarrow \beta) \vdash
  \beta
\]
in infinitary intuitionistic logic.

To verify that~(2) implies~(1), we define a semantics by the clauses in
Table~\ref{table:classifying-topos}. The following properties of the semantics
can then be established by induction:
\begin{alphlist}[(d)]
\item XXX % ...; idea: do only the propositional case
\end{alphlist}
\end{proof}

@@

\begin{table}
  \tbl{A syntactic model for XXX}{\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
    $\alpha \models \beta$ &iff& $\TT$ proves~$\alpha \vdash \beta$ \qquad(for~$\beta$ a relation symbol) \\
    $\alpha \models \top$ &iff& true \\
    $\alpha \models \varphi \wedge \psi$ &iff& $\alpha \models \varphi$ and $\alpha \models \psi$ \\
    $\alpha \models \bigvee_i \varphi_i$ &iff& there are coherent
    formulas~$\alpha_1,\ldots,\alpha_n$ such that \\ &&\qquad $\TT$ proves~$\alpha \vdash
    \alpha_1 \vee \cdots \vee \alpha_n$ and such for each~$k \in \{1,\ldots,n\}$,
    \\ &&\qquad\qquad there is an index~$i \in I$ such that~$\alpha_k \models \varphi_i$
  \end{tabular}}
  \label{table:classifying-topos}
\end{table}
